Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-05-03 ms_pre_post.Rmd
Revised manuscript with all coauthor comments

---
title: Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance
  and associated polyandry
author: ''
date: "1 May 2017"
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
  html_document: default
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from each other in real-world populations. Rather, such evolution is likely to involve multiple interacting phenotypes resulting in the potential for feedback and degeneracy (i.e., functional redundancy) among phenotypes under selection. For example, strong inbreeding depression in offspring fitness is widely hypothesised to drive selection for inbreeding avoidance, which could consequently drive selection for polyandry [e.g., @Zeh1997; @Jennions2000; @Tregenza2002]. Evolution of inbreeding avoidance can be driven by selection on both pre-copulatory and post-copulatory mechanisms, so simultaneous evolution of each phenotype might be defined by degeneracy, in addition to any direct costs on phenotypes. We used individual-based modelling to explore evolution of pre-copulatory and post-copulatory inbreeding strategy and polyandry given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given the pre-existence of the other. We thereby highlight fundamental but theoretically under-developed relationships between evolution of female mating frequency and pre-copulatory versus post-copulatory inbreeding strategy. We present our model and simulation results as a tool for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to constructing hypotheses within and across empirical systems, accurate quantitative prediction would require additional detail and the incorporation of empirical data for model parameterisation.

@@

<!--- --------------------------------------------------------------------- --->
<!--- ========== LEFT OFF HERE ============================================ --->
<!--- --------------------------------------------------------------------- --->

Evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance occurred in our model, but was affected by the joint evolution of polyandry and the existence of asymmetric costs. One cost free strategy of inbreeding avoidance precluded another more costly strategy from persisting in a focal population by modifying its effect on inbreeding avoidance (Figure 3). Consequently, long-term persistence of both pre-copulatory and post-copulatory inbreeding avoidance might not be expected in populations given sufficient cost asymmetry. Since direct costs strongly affected the evolution of the focal phenotype(s), quantifying costs is necessary for predicting evolution of inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which are also relevant for polyandry, and include costs of elevated risks of predation or disease transmission, and costs incurred through time or energy expenditure. Another relevant cost is risk of male harm caused by sexual conflict [e.g., @Arnqvist2005a; @Parker2006]. We assumed that males were passive in mating encounters with females, but inbreeding theory predicts that males should benefit from a higher tolerance of inbreeding than females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Here we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a loss of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as an up-front depletion of energy. While these costs reflect empirical observations in some populations and are therefore biologically realistic, other types of costs have also been observed and could be considered in future models tailored to specific systems. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], and such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Different types of costs might affect selection both directly and indirectly through interactions between reproductive strategies, thereby affecting evolution of entire reproductive systems. Evolution of reproductive strategies therefore needs to be considered in the broader context of potentially interacting and phenotypes and their costs.

<!--- Greta brings up a relevant point here, which is kind of in line with why I have pushed back a bit on the modelling details of the implementation over costs. I know reviewers have expressed concerns about the implementation of costs in the past, but is there any real reason to expect that our results would be much different if the cost implementation were different (assuming the *net* cost modelled were the same -- which is a fair hypothetical assumption to make because the actual values of the costs don't really mean anything except within the context of the model; i.e., you can't apply the values to empirical populations and expect to get the quantitative predictions in the model)?  --->

In our model, existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and, in turn, evolution of post-copulatory inbreeding avoidance (Figure 3C,D). However, existence of adaptive post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance (Figure 3A,B). In wild populations, it is unlikely that pre-copulatory and post-copulatory inbreeding avoidance will evolve simultaneously from an ancestral population in which females mate and assign paternity randomly. Rather, the timing of the invasion of adaptive inbreeding avoidance phenotypes will differ, so the initial evolution of one inbreeding strategy will likely occur in the absence of the other, or where selection for and subsequent evolution of the other strategy has already occurred. When framing hypotheses for existence of post-copulatory inbreeding avoidance and polyandry, it might therefore be necessary to consider whether or not inbreeding avoidance already occurs through pre-copulatory mate choice. Additionally, the opportunity for post-copulatory inbreeding avoidance will also depend on the degree to which females are polyandrous. For species in which pre-copulatory inbreeding avoidance occurs and polyandry is uncommon [@Lihoreau2007; @Metzger2010; @Metzger2010a], evolution of post-copulatory inbreeding avoidance is unlikely even if such a strategy incurs little direct cost.

@@

Post-copulatory inbreeding avoidance cannot be realised if females are not polyandrous, and is most effective for highly polyandrous females that can choose amongst multiple mates and therefore bias fertilisation accordingly. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C,D). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S9). In addition to polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, occurrence of post-copulatory inbreeding avoidance covaries positively, and pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

>>>>

---
title: Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance
  and associated polyandry
author: ''
date: "1 May 2017"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
header-includes:
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
csl: evolution.csl
bibliography: ms_pre_post.bib
---

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from each other in real-world populations. Rather, such evolution is likely to involve multiple interacting phenotypes resulting in the potential for feedback and degeneracy (i.e., functional redundancy) among phenotypes under selection. For example, strong inbreeding depression in offspring fitness is widely hypothesised to drive selection for inbreeding avoidance, which could consequently drive selection for polyandry [e.g., @Zeh1997; @Jennions2000; @Tregenza2002]. Inbreeding avoidance could be manifested through both pre-copulatory and post-copulatory mechanisms, meaning that simultaneous evolution of each phenotype might be defined by degeneracy in addition to trait-specific costs. We used individual-based modelling to highlight fundamental but theoretically under-developed relationships between evolution of polyandry and pre-copulatory versus post-copulatory inbreeding strategy given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given the pre-existence of the other. We present our model and simulation results as a tool for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to constructing hypotheses within and across empirical systems, accurate quantitative prediction would require additional detail and the incorporation of empirical data for model parameterisation.

@@

Evolution of both pre-copulatory and post-copulatory inbreeding avoidance occurred in our model, but were affected by the evolution of polyandry and the existence of asymmetric costs. One cost free strategy of inbreeding avoidance precluded another more costly strategy from persisting in a focal population (Figure 3). Consequently, long-term persistence of both pre-copulatory and post-copulatory inbreeding avoidance might not be expected in populations given sufficient cost asymmetry. Since direct costs strongly affected the evolution of the focal phenotype(s), quantifying costs is necessary for predicting evolution of inbreeding strategies.

Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which are also relevant for polyandry, and include costs of elevated risks of predation or disease transmission, and costs incurred through time or energy expenditure. Consequently, we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a loss of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as an up-front depletion of energy, all manifested as an increased probability of total reproductive failure. While these costs reflect empirical observations in some populations and are therefore biologically realistic, other types of costs have also been observed and could be considered in future models tailored to specific systems. For example, highly polyandrous females might experience increased risk of disease transmission [@Roberts2015a]. Such a cost would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005a; @Parker2006]. We assumed that males were passive in mating encounters with females, but inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could formally consider sexual conflict over both polyandry and inbreeding. Overall, such different types of costs might affect selection both directly and indirectly through interactions between reproductive strategies, thereby affecting evolution of entire reproductive systems. Evolution of reproductive strategies therefore needs to be considered in the broader context of potentially interacting and phenotypes and their costs.

In our model, existence of fixed adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and, in turn, precluded evolution of post-copulatory inbreeding avoidance (Figure 3C,D). However, existence of fixed adaptive post-copulatory inbreeding avoidance did not preclude evolution of pre-copulatory inbreeding avoidance (Figure 3A,B). In wild populations, it is unlikely that pre-copulatory and post-copulatory inbreeding avoidance will evolve simultaneously from an ancestral population in which females mate and assign paternity randomly. Rather, the timing of the invasion of adaptive inbreeding avoidance phenotypes will differ, so the initial evolution of one inbreeding strategy will likely occur in the absence of the other, or where selection for and subsequent evolution of the other strategy has already occurred. When framing hypotheses for existence of post-copulatory inbreeding avoidance and polyandry, it might therefore be necessary to consider whether or not inbreeding avoidance already occurs through pre-copulatory mate choice. Additionally, the opportunity for post-copulatory inbreeding avoidance will also depend on the degree to which females are polyandrous. For species in which pre-copulatory inbreeding avoidance occurs and polyandry is uncommon [@Lihoreau2007; @Metzger2010; @Metzger2010a], evolution of post-copulatory inbreeding avoidance is unlikely even if such a strategy incurs little direct cost.

@@

Post-copulatory inbreeding avoidance cannot be effectively realised if females are not polyandrous in any form, and is likely to be most effective for highly polyandrous females that can choose amongst sperm contributed by multiple mates and therefore bias fertilisation accordingly. In contrast, pre-copulatory inbreeding avoidance mechanisms are most critical for females that mate only once and therefore have no post-copulatory opportunity to avoiding inbreeding. This theory is borne out in our simulation results, as selection for, and consequent evolution of, post-copulatory inbreeding avoidance was negligible in populations where polyandry did not evolve, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C,D). Evolution of pre-copulatory inbreeding avoidance was also typically slightly stronger when polyandry did not evolve (e.g., Figure 3A vs 3B; see also Supporting Information Figure S9). In addition to initial polyandry causing evolution of post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back effect that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, the occurrence of post-copulatory inbreeding avoidance covaries positively, and the occurrence of pre-copulatory inbreeding avoidance covaries negatively, with the degree of polyandry. To test this hypothesis, more work is needed to quantify the degree to which females of different species engage in polyandry and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.
